---
title: "Reproducible Research - Project 1"
author: "Eduardo"
date: "8th May 2016"
output: html_document
---
# Reproducible Research: Peer Assessment 1

### Loading and preprocessing the data
```{r loaddata}
# Read activity csv file into "data" variable
data <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
```{r}
steps_by_day <- aggregate(steps ~ date, data, sum)
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")
rmean <- mean(steps_by_day$steps)
rmedian <- median(steps_by_day$steps)
```

## What is the average daily activity pattern?
```{r}
library(ggplot2)
averages <- aggregate(x=list(steps=data$steps), by=list(interval=data$interval),
                      FUN=mean, na.rm=TRUE)
ggplot(data=averages, aes(x=interval, y=steps)) +
    geom_line() +
    xlab("5-minute interval") +
    ylab("average number of steps taken")
```

On average across all the days in the dataset, the 5-minute interval contains
the maximum number of steps?
```{r}
averages[which.max(averages$steps),]
```

## Imputing missing values

There are many days/intervals where there are missing values (coded as `NA`). The presence of missing days may introduce bias into some calculations or summaries of the data.

```{r how_many_missing}
missing <- is.na(data$steps)
# How many missing
table(missing)
```

A new dataset is created that is equal to the original dataset but with the missing data filled in.

```{r}
# Replace each missing value with the mean value of its 5-minute interval
fill.value <- function(steps, interval) {
    filled <- NA
    if (!is.na(steps))
        filled <- c(steps)
    else
        filled <- (averages[averages$interval==interval, "steps"])
    return(filled)
}
filled.data <- data
filled.data$steps <- mapply(fill.value, filled.data$steps, filled.data$interval)
```
Now, using this new data set, we plot a histogram of the total number of steps taken each day and calculate the mean and median total number of steps.

```{r}
total.steps <- tapply(filled.data$steps, filled.data$date, FUN=sum)
qplot(total.steps, binwidth=1000, xlab="total number of steps taken each day")
mean(total.steps)
median(total.steps)
```

Mean and median values are higher in this new data set. That's because
in the original data, there are some days with `steps` values `NA` for 
any `interval` and this number of steps taken in such days are set to 0s by
default. 

## Are there differences in activity patterns between weekdays and weekends?
First, we are interested in knowing the day of the week for each measurement in the dataset. Let's use the dataset with the filled-in values.

```{r}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", 
              "Friday")
filled.data$dow = as.factor(ifelse(is.element(weekdays(as.Date(filled.data$date)),weekdays), "Weekday", "Weekend"))

steps_by_interval_i <- aggregate(steps ~ interval + dow, filled.data, mean)

library(lattice)

xyplot(steps_by_interval_i$steps ~ steps_by_interval_i$interval|steps_by_interval_i$dow, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")

```
